Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-08 talk/icooolps2009/paper.tex
tweaks to the benchmark section, thanks to toon again.

In this section we try to evaluate the work done so far by looking at some
benchmark numbers. Since the work is not finished, these benchmarks can only be
preliminary. Benchmarking was done on an otherwise idle machine with a 1.4
GHz Pentium M processor and 1 GB RAM, using Linux 2.6.27. All benchmarks where
run 50 times, each in a newly started process. The first run was ignored. The
final numbers were reached by computing the average of all other runs, the
confidence intervals were computed using a 95\% confidence level. All times
include the running of the tracer and machine code production.

The first round of benchmarks (Figure \ref{fig:bench1}) are timings of the
example interpreter (Figure \ref{fig:tlr-basic}) used in this paper computing
the square of 10000000 (the result will overflow, but for smaller numbers the
running time is not long enough to sensibly measure it)
using the bytecode of Figure \ref{fig:square}. The results for various
constellations are as follows:

\textbf{Benchmark 1:} The interpreter translated to C without the JIT.

\textbf{Benchmark 2:} The tracing JIT is enabled, but no inter\-preter-specific
hints are applied. This corresponds to the trace in Figure
\ref{fig:trace-normal}.  The threshold when to consider a loop to be hot is 40
iterations.  As expected, this is not faster than the previous number. It is
even quite a bit slower, probably due to the overheads of the JIT, as well as
non-optimal generated machine code.

\textbf{Benchmark 3:} The hints as in Figure \ref{fig:tlr-full} are applied, which means the loop of
the square function is reflected in the trace. Constant folding of green
variables is disabled though. This corresponds to the trace in Figure
\ref{fig:trace-no-green-folding}. This by alone brings no improvement over the
previous case.

@@

\textbf{Benchmark 5:} Same as before, but with the threshold set so high that the tracer is
never invoked. In this way the overhead of the profiling is measured. For this interpreter
it seems to be rather large, with about 20\% slowdown due to profiling. This is because the interpreter 
is small and the opcodes simple. For larger interpreters (\eg the Python one) it seems 
likely that the overhead is less significant.

@@

Furthermore, to test the technique on a more realistic example, we did some
preliminary benchmarks with PyPy's Python interpreter. The function we
benchmarked as well as the results can be seen in Figure
\ref{fig:bench-python}. The function is a bit arbitrary but executing it is
still non-trivial, as a normal Python interpreter needs to dynamically dispatch
nearly all of the involved operations (like indexing into the tuple, addition
and comparison of \texttt{i}). We benchmarked PyPy's Python interpreter with the
JIT disabled, with the JIT enabled and
CPython\footnote{\texttt{http://python.org}} 2.5.4 (the reference implementation of
Python).

>>>>

In this section we evaluate the work done so far by looking at some
benchmark numbers. Since the work is not finished, these benchmarks can only be
preliminary. Benchmarking was done on an otherwise idle machine with a 1.4
GHz Pentium M processor and 1 GB RAM, using Linux 2.6.27. All benchmarks where
run 50 times, each in a newly started process. The first run was ignored. The
final numbers were reached by computing the average of all other runs, the
confidence intervals were computed using a 95\% confidence level. All times
include running the tracer and producing machine code.

The first round of benchmarks (Figure \ref{fig:bench1}) are timings of the
example interpreter given in Figure \ref{fig:tlr-basic} computing
the square of 10000000\footnote{the result will overflow, but for smaller numbers the
running time is not long enough to sensibly measure it}
using the bytecode of Figure \ref{fig:square}. The results for various
configurations are as follows:

\textbf{Benchmark 1:} The interpreter translated to C without including a JIT
compiler.

\textbf{Benchmark 2:} The tracing JIT is enabled, but no inter\-preter-specific
hints are applied. This corresponds to the trace in Figure
\ref{fig:trace-normal}.  The threshold when to consider a loop to be hot is 40
iterations.  As expected, this is not faster than the previous number. It is
even quite a bit slower, probably due to the overheads, as well as
non-optimal generated machine code.

\textbf{Benchmark 3:} The tracing JIT is enabled and hints as in Figure
\ref{fig:tlr-full} are applied. This means that the interpreter loop is unrolled
so that it corresponds to the loop in the square function. However, constant folding of green
variables is disabled, therefore the resulting machine code corresponds to the
trace in Figure \ref{fig:trace-no-green-folding}. This by alone brings no
improvement over the previous case.

@@

\textbf{Benchmark 5:} Same as before, but with the threshold set so high that the tracer is
never invoked. In this way the overhead of the profiling is measured. For this interpreter
it seems to be rather large, with about 20\% slowdown due to profiling. This is because the interpreter 
is small and the opcodes simple. For larger interpreters (\eg PyPy's Python
interpreter) the overhead will likely be less significant.

@@

To test the technique on a more realistic example, we did some
preliminary benchmarks with PyPy's Python interpreter. The function we
benchmarked as well as the results can be seen in Figure
\ref{fig:bench-python}. The function is a bit arbitrary but executing it is
still non-trivial, as a normal Python interpreter needs to dynamically dispatch
nearly all of the involved operations (like indexing into the tuple, addition
and comparison of \texttt{i}). We benchmarked PyPy's Python interpreter with the
JIT disabled, with the JIT enabled and
CPython\footnote{\texttt{http://python.org}} 2.5.4 (the reference implementation of
Python).
